Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-04 Manuscript/manuscript.tex
Manuscript small update to Samuli's writing

In this work, we empirically sample this parameter space $f_q \in [0.75, 1.0]$
to find an atomistic MD simulation force field parameters for lipids that
accurately describe the lipid head group response to varying concentrations of cations
when compared to NMR data \cite{akutsu81,altenbach84,scherer89}. Such data can be used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, applied to headgroup, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related headgroup order parameter responses become weaker. 
The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and 
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of bound charge in a lipid bilayer \cite{seelig87}.
This is rationalized as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
More recent analysis in the NMRlipids project concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. The concept was also shown
to be suitable for the comparison of ion binding affinity in lipid 
bilayers between simulations and experiments~\cite{catte16}.

The order parameters for $\alpha$ and $\beta$ segments in headgroup, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using $^2$H NMR or $^{13}$C NMR techniques.
The order parameters can be calculated from MD simulations 
by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Before using electrometer concept to analyze cation binding, 
we calibrate the response of headgroup order parameters to the bound
charge in the simulation models. This is done by comparing 
the slope $m_i$ in Eq. \ref{OPchangeEQ} between simulations and
experiments of cationic surfactants in POPC bilayer \cite{scherer89}. 
Since essentially all surfactants locate in bilayer, the 
molar ratio directly gives the amount of bound charge per 
lipid $X^\pm$ is these systems.

\subsection{Ion concentrations and binding affinity}
The experimental headgroup order parameter data used
in this work is reported as a function of 
cation concentration remaining in the bulk 
solvent \cite{akutsu81,altenbach84}. 
To determine the same concentration from simulations, we
use large enough box to ensure that the ion concetration
reaches a clear plateau and the electroneutrality
is satisfied at the edge of simulation box. It should be noted that the recent work
by NMRlipids project compared simulations to the same 
experimental data by using the relative molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M \cite{catte16}.
This is not exactly the same as the concentration reported in the used experimental
data, because the binding of ions in bilayer reduces the bulk concentration.
The difference slightly moves points towards lower concentrations but do not 
affect the conclusions in the previous study \cite{catte16}.

The ion binding affinities between different models were compared by calculating
the relative surface excess of ions with respect to water, $\Gamma_i^w$, as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that phospholipid membranes are bilayers, 
and hence the interface occurs in the simulation box twice
providing a surface area $2A$.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, can be determined
from the concentration plateau at the edge of the simulation box.

@@

The order parameters were calculated from simulations for all C-H bonds
in lipid molecule by using Eq. \ref{OP}. Form factors were calculated 
from equation
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
The systems were simulated for 200~ns using Lipid14 model for POPC.
First 20~ns was omitted from the analysis.

The same systems were also simulated with ECC-lipid model for POPC \todo{How long?}.
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all \todo{Were all scaled? Are there charges is tails? Were those scaled as well in this case?} 
charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\subsection{Response of POPC headgroup to bound charge}

@@

Before proceeding to ion binding affinity studies, we quantify
the response of headgroup order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is 
given by the molar fraction of cationic surfactants, because essentially all surfactants 
locate in the lipid bilayers. Comparison of headgroup order parameter
changes between simulations and experiments 
as a function of mole fraction of cationic surfactants 
can be then used to validate the sensitivity of
lipid headgroup to bound charge in simulation models.

%depends on both, the
%binding affinity and the response of the headgroup on bound charge.
%Thus, the headgroup response to bound charge has to be correctly
%described in the model if the concept is used for quantitative studies of
%binding affinity. Hence, we first quantify the response of head group order
%parameters to the bound charge by using experimental data measured from
%mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.
%
%Thus, the headgroup order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The headgroup order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

The headgroup order parameter changes as a function of mole fraction
of cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease of the order parameters is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the headgroup order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

Also the effect of bound cations on the
headgroup P-N vector angle with respect to the membrane normal
is shown in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more parallel to the membrane
normal with an increasing amount of bound cations, as expected~\cite{seelig87}. 
The orientation effect is more pronounced in Lipid14 model than 
in ECC-lipids model, which is in line with the order parameter 
results and reduced charge--dipole interactions in the ECC-lipid model.
Since the response of $\alpha$-order parameter is in very good agreement
with experiments in ECC-lipid model, we can suggest a relation between
order parameter and P-N vector angle changes. The data gives
$\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}$, which can be used
to estimate P-N vector angle changes from experimental results
for $\alpha$ order parameters. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Changes of headgroup order parameters of POPC bilayer as a function of NaCl and CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the plateau of cation number density $C_{np}$
    at the box edge as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
\end{figure*}

The linear decrease of the headgroup order parameters, quantified in previous 
section, can be used to measure the amount of bound aqueous cations in lipid 
bilayers~\cite{seelig87,catte16}. The headgroup order parameter response 
to NaCl and CaCl$_2$ concentrations from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
%are shown together with the results from Lipid14 simulation
%with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

As already discussed previously~\cite{catte16}, NaCl has a negligible effect 
on the headgroup order parameters with submolar 
concentrations due to the very weak binding in PC bilayers. 
In contrast to almost all the other simulation models, this behaviour 
is reproduced by Lipid14 model with \AA{}qvist ions \cite{catte16}. 
The headgroup order parameter response to CaCl$_2$ concentration was,
however, overestimated with the same combination of force field
parameters~\cite{catte16}. The results were not improved when
ion models by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour were used, as seen in 
in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
Together with extensive data in previous work \cite{catte16}, 
the results suggest that CaCl$_2$ binding behaviour in PC bilayers
cannot be corrected by only improving the ion models.

>>>>

In this work, we empirically sample this parameter space $f_q \in [0.75, 1.0]$
to find an atomistic MD simulation force field parameters for lipids that
accurately describe the lipid head group response to varying concentrations of cations
when compared to NMR data \cite{akutsu81,altenbach84,scherer89}. Such data can be used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and head group
parameters are required in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, applied to head group, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related head group order parameter responses become weaker. 
The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the decreased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
This increases solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Comparing head group order parameter response to bound charge in simulations and experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and 
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of bound charge in a lipid bilayer \cite{seelig87}.
This is rationalized as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
More recent analysis in the NMRlipids project concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. The concept was also shown
to be suitable for the comparison of ion binding affinity in lipid 
bilayers between simulations and experiments~\cite{catte16}.

The order parameters for $\alpha$ and $\beta$ segments in head group, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using $^2$H NMR or $^{13}$C NMR techniques.
The order parameters can be calculated from MD simulations 
by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Before using electrometer concept to analyze cation binding, 
we calibrate the response of head group order parameters to the bound
charge in the simulation models. This is done by comparing 
the slope $m_i$ in Eq. \ref{OPchangeEQ} between simulations and
experiments of cationic surfactants in POPC bilayer \cite{scherer89}. 
Since essentially all surfactants locate in bilayer, the 
molar ratio directly gives the amount of bound charge per 
lipid $X^\pm$ is these systems.

\subsection{Ion concentrations and binding affinity}
The experimental head group order parameter data used
in this work is reported as a function of 
cation concentration remaining in the bulk 
solvent \cite{akutsu81,altenbach84}. 
\todo{JOE: Following is hard to understand 
(even though I think I know the idea. 
Should be rewritten so that it is clear.}
To determine the same concentration from simulations, we
use large enough box to ensure that the ion concetration
reaches a clear plateau and the electroneutrality
is satisfied at the edge of simulation box. It should be noted that the recent work
by NMRlipids project compared simulations to the same 
experimental data by using the relative molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M \cite{catte16}.
This is not exactly the same as the concentration reported in the used experimental
data, because the binding of ions in bilayer reduces the bulk concentration.
The difference slightly moves points towards lower concentrations but do not 
affect the conclusions in the previous study \cite{catte16}.

The ion binding affinities between different models were compared by calculating
the relative surface excess of ions with respect to water, $\Gamma_i^w$, as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that phospholipid membranes are bilayers, 
and hence the interface occurs in the simulation box twice
providing a surface area $2A$.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, can be determined
\todo{JOE: I wouldn't use the term plateau. Term ``region where concentration is constant'' is preferred. Applies to the whole text -- replace everywhere. }
from the concentration plateau at the edge of the simulation box.

@@

The order parameters were calculated from simulations for all C-H bonds
in lipid molecules by using Eq. \ref{OP}. Form factors were calculated 
from equation
\begin{equation}
  F(q) = \int _{-D/2} ^{D/2} \left ( \sum _\alpha f_\alpha (q_z) n_\alpha (z) - \rho _s \right ) \exp (izq_z) \mathrm{d}z,
\end{equation}
where $f_\alpha(q_z)$ is the density of atomic scattering length, 
$\rho_s$ is the density of solvent scattering length,
$n_\alpha (z)$ is the number density of atom $\alpha$ and
$z$ is the distance from the membrane centre along its normal 
spanning the membrane with thickness $D$.

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
The systems were simulated for 200~ns using Lipid14 model for POPC.
First 20~ns were omitted from the analysis.

The same systems were also simulated with ECC-lipid model for POPC using the same setup. 
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids. 
\todo{Were all scaled? Are there charges is tails? Were those scaled as well in this case? 
      -- JOE: yes, all scaled for simplicity. }

@@

\subsection{Response of POPC head group to bound charge}

@@

Before proceeding to ion binding affinity studies, we quantify
the response of head group order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is 
given by the molar fraction of cationic surfactants, because essentially all surfactants 
locate in the lipid bilayers. Comparison of head group order parameter
changes between simulations and experiments 
as a function of mole fraction of cationic surfactants 
can be then used to validate the sensitivity of
lipid head group to bound charge in simulation models.

%depends on both, the
%binding affinity and the response of the head group on bound charge.
%Thus, the head group response to bound charge has to be correctly
%described in the model if the concept is used for quantitative studies of
%binding affinity. Hence, we first quantify the response of head group order
%parameters to the bound charge by using experimental data measured from
%mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.
%
%Thus, the head group order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The head group order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

The head group order parameter changes as a function of mole fraction
of cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease of the order parameters is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

Also the effect of bound cations on the
head group P-N vector angle with respect to the membrane normal
is shown in Fig.~\ref{OrderParameterCHANGESsurf}.
The head group orients more parallel to the membrane
normal with an increasing amount of bound cations, as expected~\cite{seelig87}. 
The orientation effect is more pronounced in Lipid14 model than 
in ECC-lipids model, which is in line with the higher order parameter 
response and reduced charge--dipole interactions in the ECC-lipid model.
Since the response of $\alpha$-order parameter is in very good agreement
with experiments in ECC-lipid model, we can suggest a relation between
order parameter and P-N vector angle changes. The data gives
$\Delta \Theta_{\rm P-N}=?? \Delta S_{\rm CH}$, which can be used
to estimate P-N vector angle changes from experimental results
for $\alpha$ order parameters. 
\todo{SAMULI: I think that we should establish the relation between 
order parameter and P-N vector angle changes by using ECC-lipids model.
This would be useful for people measuring aplha order parameter
changes from experiments.}

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Changes of head group order parameters of POPC bilayer as a function of NaCl and CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the plateau of cation number density $C_{np}$
    at the box edge as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
\end{figure*}

The linear decrease of the head group order parameters, quantified in previous 
section, can be used to measure the amount of bound aqueous cations in lipid 
bilayers~\cite{seelig87,catte16}. The head group order parameter response 
to NaCl and CaCl$_2$ concentrations from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
%are shown together with the results from Lipid14 simulation
%with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.
